RNApyro

### 2012-10-05 Recomb/methods_RECOMB.tex
method

%!TEX root = main_RECOMB.tex
\section{Methods}
\label{sec:methods}
Let $\Omega$ be an un-gapped RNA alignment, $S$ its associated secondary structure, $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the SCFG generating  
all  secondary structures of length  $|s|$. We are interested in  
the probability of a given position being a specific nucleotides,
over all sequences having at most $m$ mutations from $s$, under the SCFG derivation $S$ 
(formally 
$\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define as a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former, as the 
equivalent of the \emph{inside}, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing $m$ mutations. The latter computes the \emph{outside}, in particular,  
the partition function considering only between $[0,i]\cup[j,n-1]$
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing 
 $m$ mutations outside.

@@

\[
	\forall i \in (0,\cdots,n-1):\, \Z{i+1,i}{m}{a,b}=\left\{
	\begin{array}{ll}
		1 &\text{If } m = 0\\
		0 &\text{Else }
	\end{array}\right.
\]
In other words, when we evaluate the function $\mathcal Z$, after exhausting all positions, there 
is only one possible solution if there is $0$ mutations left, and none else. Since the 
energetic terms only depend on base pairs, they are not involved in the initial conditions.

The recursion itself is composed of four terms:
$$
	\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} & \text{If }S_{i}=-1\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-(\alpha \text{ES}^\beta_{a b \to a' b'}+(1-\alpha)\text{EI}^{\Omega}_{(i,j),a'b'})}{RT}}
			 \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}&
			 \text{Elif }S_i=j \land S_{i-1}=j+1\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(i,k),a'b'}}{RT}}
      \Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \Z{k+1,j}{m'}{b',b} & \text{Elif }S_i=k \land i < k \leq j\\
      0 &\text{Else}
	\end{array}\right.
$$
The cases can be broked down as follows.
\begin{description}
\item[$S_{i}=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we increase the lower interval bound by $1$ (i.e. $i+1$), and consider all possible
 nucleotides $a'$ at position $i$, updating the number of mutants in function of $\Kron_{a',s_i}$. 
\item[$S_i=j$ and $S_{i-1}=j+1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i-1$ is
paired with $j+1$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{inside} value of the interval $[i+1,j-1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_i=k$ and $i<k \leq j$:] If nucleotide $i$ is paired with position $k$ 
but is not stacked outside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. This 
creates
two different intervals for which we must compute the values, $[i+1,k-1]$ and $[k+1,j-1]$, for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $j$ respectively.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.
\end{description}

\subsection{Outside}	
The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
energy in subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
(resp. in position $j-1$ it is $b$).
We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
$$
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
$$
The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
 Thus, when we need
to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
structure between $(j,n-1)$ must be a set of independent stems. Therefore,
 all the outside energy between $[j,n-1]$ is
equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself is as follows.
$$
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
    \text{Elif }S_i=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
		 e^{\frac{-(\alpha \text{ES}^\beta_{a b \to a' b'}+(1-\alpha)\text{EI}^{\Omega}_{(i,j),a'b'})}{RT}}
    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(i,k),a'b'}}{RT}}
		 \Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S_{i}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(k,i),a'b'}}{RT}}
		 \Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S_{i}=k < i\\
		 0 & \text{Else}
  \end{array}\right.
$$
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{forward}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

@@

We are now interested in knowing, under our model, 
 the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of mutations allowed. 
$$
	\mathbb{P}(s_i = x\mid s,\Omega, S,M):=\left\{
	\begin{array}{ll}
		\displaystyle
		\frac{
			\displaystyle
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		}{
			\displaystyle
			\sum_{m=0}^{M}
			\Z{0,n-1}{m}{X,X}
		}
		&\text{If }S_i = -1\\
		\displaystyle
		\frac{
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in Bases\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(i,k),xb}}{RT}}
			\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\Z{i+1,k-1}{m'}{x,b}
		}{
			\displaystyle
			\sum_{m=0}^{M}
			\Z{0,n-1}{m}{X,X}
		}
		&\text{If }S_i=k>i\\
		\displaystyle
 		\frac{
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in Bases\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(k,i),bx}}{RT}}
			\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\Z{k+1,i-1}{m'}{b,x}
		}{
			\displaystyle
			\sum_{m=0}^{M}
			\Z{0,n-1}{m}{X,X}
		}
		&\text{If }S_i=k<i
	\end{array}\right.
$$

>>>>

%!TEX root = main_RECOMB.tex
\section{Methods}
\label{sec:methods}
Let $\Omega$ be an un-gapped RNA alignment, $S$ its associated secondary structure, $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the SCFG generating  
all  secondary structures of length  $|s|$. We are interested in  
the probability of a given position being a specific nucleotides,
over all sequences having at most $m$ mutations from $s$, under the SCFG derivation $S$ 
(formally 
$\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define as a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former presented in 
Eq.~\ref{eq:Z_in} and Eq.~\ref{eq:Z_rec}, our 
version of the \emph{inside}, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing 
$m$ mutations. The latter in Eq.~\ref{eq:Y_in} and Eq.\ref{eq:Y_rec},
 computes the \emph{outside}, in particular,  
the partition function considering only between $[0,i]\cup[j,n-1]$
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing 
 $m$ mutations outside.

@@

\begin{equation}
	\forall i \in (0,\cdots,n-1):\, \Z{i+1,i}{m}{a,b}=\left\{
	\begin{array}{ll}
		1 &\text{If } m = 0\\
		0 &\text{Else }
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, when we evaluate the function $\mathcal Z$, after exhausting all positions, there 
is only one possible solution if there is $0$ mutations left, and none else. Since the 
energetic terms only depend on base pairs, they are not involved in the initial conditions.

The recursion itself is composed of four terms:
\begin{equation}
	\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} & \text{If }S_{i}=-1\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-(\alpha \text{ES}^\beta_{a b \to a' b'}+(1-\alpha)\text{EI}^{\Omega}_{(i,j),a'b'})}{RT}}
			 \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}&
			 \text{Elif }S_i=j \land S_{i-1}=j+1\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(i,k),a'b'}}{RT}}
      \Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \Z{k+1,j}{m'}{b',b} & \text{Elif }S_i=k \land i < k \leq j\\
      0 &\text{Else}
	\end{array}\right.
\label{eq:Z_rec}
\end{equation}
The cases can be broked down as follows.
\begin{description}
\item[$S_{i}=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we increase the lower interval bound by $1$ (i.e. $i+1$), and consider all possible
 nucleotides $a'$ at position $i$, updating the number of mutants in function of $\Kron_{a',s_i}$. 
\item[$S_i=j$ and $S_{i-1}=j+1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i-1$ is
paired with $j+1$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{inside} value of the interval $[i+1,j-1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_i=k$ and $i<k \leq j$:] If nucleotide $i$ is paired with position $k$ 
but is not stacked outside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. This 
creates
two different intervals for which we must compute the values, $[i+1,k-1]$ and $[k+1,j-1]$, for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $j$ respectively.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.
\end{description}

\subsection{Outside}	
The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
energy in subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
(resp. in position $j-1$ it is $b$).
We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
\begin{equation}
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
\label{eq:Y_in}
\end{equation}
The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
 Thus, when we need
to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
structure between $(j,n-1)$ must be a set of independent stems. Therefore,
 all the outside energy between $[j,n-1]$ is
equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself is as follows.
\begin{equation}
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
    \text{Elif }S_i=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
		 e^{\frac{-(\alpha \text{ES}^\beta_{a b \to a' b'}+(1-\alpha)\text{EI}^{\Omega}_{(i,j),a'b'})}{RT}}
    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(i,k),a'b'}}{RT}}
		 \Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S_{i}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(k,i),a'b'}}{RT}}
		 \Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S_{i}=k < i\\
		 0 & \text{Else}
  \end{array}\right.
\label{eq:Y_rec}
\end{equation}
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{forward}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

@@

We are now interested in knowing, under our model, 
 the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of mutations allowed. 
\begin{equation}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M):=\left\{
	\begin{array}{ll}
		\displaystyle
		\frac{
			\displaystyle
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		}{
			\displaystyle
			\sum_{m=0}^{M}
			\Z{0,n-1}{m}{X,X}
		}
		&\text{If }S_i = -1\\
		\displaystyle
		\frac{
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in Bases\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(i,k),xb}}{RT}}
			\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\Z{i+1,k-1}{m'}{x,b}
		}{
			\displaystyle
			\sum_{m=0}^{M}
			\Z{0,n-1}{m}{X,X}
		}
		&\text{If }S_i=k>i\\
		\displaystyle
 		\frac{
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in Bases\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-(1-\alpha)\text{EI}^{\Omega}_{(k,i),bx}}{RT}}
			\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\Z{k+1,i-1}{m'}{b,x}
		}{
			\displaystyle
			\sum_{m=0}^{M}
			\Z{0,n-1}{m}{X,X}
		}
		&\text{If }S_i=k<i
	\end{array}\right.
\end{equation}
